---
title: ""
output: pdf_document
---

The dataset spam available in package kernlab contains 57 features extracted from the content of emails which were classified as spam or nonspam. The first 48 variables contain the frequency of the variable name (e.g., business) in the email. If the variable name starts with num (e.g., num650) it indicates the frequency of the corresponding number (e.g., 650). The variables 49-54 indicate the frequency of the special characters ;, (, [, !, \$, and \#. The variables 55-57 contain the average, longest and total run-length of capital letters. The last variable, type indicates if an email is spam or nonspam.

The task is to predict if an email is spam or not, using columns 49-57.

#   Prepare data
```{r}
# Load the data
data("spam", package = "kernlab")

# Select relevant columns
data_spam = spam[, 49:58]

# check if all variables are coded correctly (variable type)
str(data_spam)
```

#   Logistic Regression
```{r}
spam_model = glm(type ~ ., data = data_spam, family = binomial)
summary(spam_model)
```

#   Predictive performance - Accuracy and ROC
```{r}
# Set tau
tau = 0.5
predicted_p = fitted(spam_model)
y_hat = ifelse(predicted_p > tau, 1, 0)

# Cross tabulation between observed and predicted
table(data_spam$type, y_hat)

# Compute accuracy for given tau
tab = table(data_spam$type, y_hat)
sum(diag(tab)) / sum(tab)
```
The larger the area under the ROC curve the better.
The advantage of using the ROC and AU-ROC is that the predictive performance is evaluated regardless of a specific choice of τ.
```{r}
# Package ROCR can be used to calculate many performance measures
# Varying Tau will give different values of various performance measures

library(ROCR)

# Create a prediction object, stores predicted probabilities and true class labels
pred_object = prediction(predicted_p, data_spam$type)

# Plot ROC Curve
# Function performance then can be applied to calculate different performance measure
# ?performance to see all metrics available
roc = performance(pred_object, "tpr", "fpr")
plot(roc)
abline(0, 1, col = "darkorange2", lty = 2) # add bisect line
```

```{r}
# compute the area under the ROC curve
auc = performance(pred_object, "auc")
auc@y.values
```
#   Predictive performance - Sensitivity(TPR) and Specificity(TNR)
```{r}
# Sensitivity TPR
tpr = performance(pred_object, "tpr")

# Specificity TNR
tnr = performance(pred_object, "tnr")
```

The optimal threshold𝜏can be found maximizing the sum of TPR and TNR
```{r}
# values of tau from 0 to 1
tau = tpr@x.values[[1]]
# x.values are the same for all metrics 0 - 1


tpr_tnr = tpr@y.values[[1]] + tnr@y.values[[1]]
best_roc = which.max(tpr_tnr)

plot(tau, tpr_tnr, type = "l", ylab = "tpr + tnr")
points(tau[best_roc], tpr_tnr[best_roc], pch = 19, col = adjustcolor("darkorange2", 0.5))
```
```{r}
tau[best_roc] # optimal tau
```
```{r}
# Classification using optimal tau
pred_y = ifelse(fitted(spam_model) > tau[best_roc], 1, 0)
table(data_spam$type, pred_y)
```
```{r}
# accuracy for optimal tau
acc = performance(pred_object, "acc")
acc@y.values[[1]][best_roc]
```
```{r}
# sensitivity and specificity for optimal tau
tpr@y.values[[1]][best_roc]
tnr@y.values[[1]][best_roc]
```

#   Predictive performance - Precision/Recall/F1 score
```{r}
# produce precision/recall curve
pr = performance(pred_object, "prec", "rec")
plot(pr)
```
```{r}
# compute area under the PR curve
aucpr = performance(pred_object, "aucpr")
aucpr@y.values # add this to graph
# the higher the better AUC-PR
```
f1 score: The score can be interpreted as a model’s balanced ability to both detect positive cases (recall) and be accurate with the cases it detects (precision). The F1 score is the harmonic mean of precision and recall:
Pick the threshold with the highest f1 score if focus is predicting the positive cases

```{r}
f1 = performance(pred_object, "f")
f1_scores = f1@y.values[[1]]
best_f1 = which.max(f1_scores)
```

```{r}
plot(tau, f1_scores, type = "l")
points(tau[best_f1], f1_scores[best_f1], pch = 19, col = adjustcolor("darkorange2", 0.5))

tau[best_f1] # optimal tau
f1_scores[best_f1] # best f1
acc@y.values[[1]][best_f1] # accuracy using optimal tau as threshold
```
If the classes are significantly imbalanced F1 and Sens+Spec will result in different optimal tau



#   Multinomial logistic regression: framework for multi-class classification problems

```{r}
library(nnet)
data("Vehicle", package = "mlbench")
vehicle_data = Vehicle
```

```{r}
vehicle_model = multinom(Class ~ ., data = vehicle_data, maxit = 300)
table(vehicle_data$Class)
summary(vehicle_model)
```


```{r}
table(predict(vehicle_model)) # by default gives predicted classes instead of probability
```


```{r}
tab = table(true = vehicle_data$Class, pred = predict(vehicle_model))
tab
```


```{r}
# Accuracy
sum(diag(tab)) / sum(tab)
```


```{r}
# Sensitivity TPR
cbind(tab, sens = diag(tab) / rowSums(tab))
```


```{r}
# 2.1 TASK
data("Vowel", package = "mlbench")
vowel_data = Vowel

vowel_model = multinom(Class ~ ., data = vowel_data, maxit = 500)

# true classes
table(vowel_data$Class)

# predicted classes
table(predict(vowel_model))

tab = table(true = vowel_data$Class, pred = predict(vowel_model))
tab

acc = sum(diag(tab)) / sum(tab)
acc

# tpr
diag(tab) / rowSums(tab)
```



```{r}

heart_data = read.csv("data_heart_disease_BRFSS2015.csv")

#factor binary/categorical variables
heart_data[, c(1:4, 6:22)] = lapply(heart_data[, c(1:4, 6:22)], factor)
str(heart_data)

# Classes highly imbalanced
y = table(heart_data$HeartDiseaseorAttack)
table(heart_data$HeartDiseaseorAttack) / nrow(heart_data)

tau = 0.5

heart_model = glm(HeartDiseaseorAttack ~ ., data = heart_data, family = binomial, maxit = 1000) 
heart_model

# Predicted
predicted_p = fitted(heart_model)
y_hat = ifelse(predicted_p > tau, 1, 0)

table(y_hat)





data <- read.csv("data_heart_disease_BRFSS2015.csv")
# make sure that binary/categorical variables are correctly encoded as factor
data[,c(1:4,6:14,18:19)] <- lapply( data[,c(1:4,6:14,18:19)], factor )
str(data)
# classes are highly imbalanced
table(data$HeartDiseaseorAttack)
table(data$HeartDiseaseorAttack)/nrow(data)


y = data$HeartDiseaseorAttack
table(data$HeartDiseaseorAttack) / nrow(data)

heart_model = glm(HeartDiseaseorAttack ~ ., data = data, family = binomial, maxit = 1000) 
heart_model

predicted_p = fitted(heart_model)
y_hat = ifelse(predicted_p > tau, 1, 0)
table(y_hat)

tab = table(true = y, pred = y_hat)
tab

# accuracy
sum(diag(tab)) / sum(tab)

pred_object = prediction(predicted_p, data$HeartDiseaseorAttack)


# Sensitivity TPR
tpr = performance(pred_object, "tpr")

# Specificity TNR
tnr = performance(pred_object, "tnr")

# the optimal threshold𝜏can be found maximizing the sum of TPR and TNR

# values of tau from 0 to 1
tau = tpr@x.values[[1]]
# x.values are the same for all metrics 0 - 1

tpr_tnr = tpr@y.values[[1]] + tnr@y.values[[1]]
best_tau = which.max(tpr_tnr)
tau[best_tau]

plot(tau, tpr_tnr, type = "l", ylab = "tpr + tnr")
points(tau[best_tau], tpr_tnr[best_tau], color = "red")

# produce precision/recall curve
pr = performance(pred_object, "prec", "rec")
plot(pr)


f1 = performance(pred_object, "f")
f1_values = f1@y.values[[1]]
best_f1 = which.max(f1_values)
best_f1
tau = f1@x.values
tau[best_f1]
plot(tau, f1_values)
```



